---
title: "Investigating the Relationship Between Asthma and Pollution Burden in California to Understand the Importance of Race and Economic Status as Social Determinants of Health"
author: "Maggie Walsh"
date: "May 5th, 2020"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, error = FALSE, working = FALSE)
```

# Introduction

Air pollution has become an increasingly important topic of research in the past half-century, mainly because of the multitude of health risks that have been shown to accompany exposure to air pollution such as respiratory disease, cardiovascular disease, and even neurological dysfunction [@brunekreef2002air; @rajagopalan2018air; @schikowski2020role]. Air pollution consists of a multitude of substances from a multitude of sources. Air pollution arises from fuel burning for energy and transportation, and as a product of other industrial processes. Pollution from mobile sources such as motor traffic has been cited as one of the most prevalent sources of harmful air pollution, especially in cities, where a vast population uses cars to commute into and out of dense urban areas daily [@mayer1999air].

Asthma is a respiratory disease that has been linked to air pollution in numerous studies worldwide [@koenig1999air; @raji2020acute]. There are studies that look specifically at the impact of air pollution from mobile sources on the development and worsening of asthma  [@gasana2012motor; @ristovski2012respiratory]. These studies investigate a number of different components of motor traffic pollution which contribute to asthma including particulate matter. Particulate matter less than 2.5 micrometers in diameter, known as fine particulate matter, can have a hazardous effect because it is small enough to be inhaled into small airways and alveoli (small air sacs) in the lung [@guarnieri2014outdoor]. These fine particles consist of transition metals and other substances that can cause the symptoms associated with asthma attacks [@guarnieri2014outdoor]. These pollutants, which emerge from cars on major roadways in dangerous concentrations, can have these effects up to 500 meters away [@guarnieri2014outdoor]. For reference, that is the distance covered by five and a half football fields. In North American cities, 30-45% of people live within this distance of a major roadway [@guarnieri2014outdoor].

Air pollution from motor traffic has a particularly pronounced impact in the state of California. In the American Lung Association's State of the Air Report for 2019, 6 out of 10 of the countries most polluted cities for year round particle pollution were California cities. California relies heavily on car commuting because of the development of cities and surrounding infrastructure in a way that prioritized travel by motor vehicles [@cervero2003growing]. The problem is currently worsening because rapidly growing urban areas in Southern California and the Bay Area promise new jobs, but no affordable housing opportunities [@cervero2003growing].Because of the disparity between job opportunities and affordable housing, Californians are forced to drive long distances to work and thereby increase the mobile source pollution in these metropolitan areas. This may contribute to the placement of the Los Angeles and San Jose/San Francisco metropolitan areas on the top ten list for most particle polluted cities.

The problem of air pollution and asthma does not impact all Californians equally. Differences in race and socioeconomic status has been shown to relate to how much particle pollution from motor traffic one is exposed to [@houston2004structural; @houston2014disparities]. Race and socioeconomic status, as they relate to potential health impacts, have come to be known as social determinants of health [@marmot2005social]. These social determinants of health reflect how racism and class-ism extend beyond individual viewpoints, into the fabric of our society.

In this study, I aimed to highlight this relationship between asthma and traffic pollution in California Counties using asthma emergency department visit rates and pollution data provided by the California Enviroscreen. Then, I questioned the relationship between poverty, particulate matter pollution, and asthma in San Francisco. I chose to focus on San Francisco because in California, much of the focus on the health impact of traffic pollution, as well as the social justice implications, have been based on studies conducted in Southern California [@houston2004structural; @houston2014disparities; @perez2009global]. San Francisco and the surrounding cities of the bay area face the same problems with job demand and a lack of affordable housing. There was a strong relationship between Particulate matter from diesel vehicles and asthma in San Francisco There is also evidence that these same neighborhoods with higher diesel particulate matter and higher asthma rates also had higher poverty. I identified a strong relationship between asthma and poverty in the San Francisco region. With a consideration of race as a social determinant of health, I used data from the California Department of Public Health to investigate the relationship between asthma emergency department visit rates and race in California. This inquiry revealed a striking increase in asthma emergency department visits in black Californians state wide.

# Methods

## Data Sources

Data was taken from the California Office of Environmental Health Hazard Assessment Website (Enviroscreen) and the California Department of Public Health Website (asthma emergency department visits by county) Data from the California Enviroscreen were used in the analysis. The California Enviroscreen is a comprehensive data set collecting data from each California census tract regarding different indicators calculated using data from a variety of sources. I used several of these indicators in my analysis.

One indicator was from a group of "sensitive population indicators", revealing populations of individuals with certain biological traits which make them especially vulnerable to the impacts of pollution. Asthma is an indicator of sensitivity to air pollution. The age adjusted rates of emergency department visits due to asthma were given for each California census tract to indicate the presence of such sensitive populations. Each emergency department visit to hospitals licensed by California is required to be reported to the California Office of Environmental Health Hazard Assessment. This report includes the principal diagnosis, which was used in this data set to calculate the average rates of asthma visits per 10,000 residents between the years of 2011 and 2013. 
"Pollution Burden" indicators such as the pollution burden score given by the Enviroscreen to the County census tracts was also used in my analysis. The pollution burden score reported by the California Enviroscreen was determined by taking the average percentiles of the seven "exposures indicators" (ozone concentration, PM 2.5 concentration, diesel PM emissions, drinking water contaminants) and the five environmental "effects indicators" (cleanup sites, impaired water bodies, groundwater threats, hazardous waste facilities and generators, and solid waste sites and facilities). The effects indicators were given half the weight of the exposures indicators.This calculated average of the indicators was divided by ten for a burden score ranging from .1 to 10

Another indicator relevant to my analysis was mean fine particulate matter (PM 2.5). The mean PM 2.5 concentrations were estimated using data from monitoring sites in the California Air Resources Board monitoring network. All data used was from the years 2012 to 2014. The data represents an estimate of the mean concentration at the center of each census tract location. The annual mean was calculated using the quarterly estimates then averaged across the three years. The values for census tracts with geographical centers over 50 kilometers away from the nearest monitor were given a concentration based on satellite observations from the years between 2006 and 2012.

A similar indicator used in my analysis was diesel particulate matter (Diesel PM). This value represented a gridded distribution of county emission averages from road and non road sources for a 2012 July weekday day. This data was taken from the California Air Resources Board using the on road emissions model, EMFAC2013, for on-road sources and the emissions inventory forecasting system, CEPAM, for off road sources. the results were summed into a gridded data set. These estimates were then allocated to the various California Census Tracts using Arc Map.

An important "Socioeconomic Factor Indicator," poverty, was used in my analysis as well. The Enviroscreen represented poverty levels in each of the California Census tracts by taking data from the 2011-2015 American Community Survey that included the number of people below 200 percent of the federal poverty level. The number of people under this line were divided by the sample size to get an estimated percent poverty level. The standard error and relative standard error were calculated to make sure these estimates were reliable.

The data used in my analysis regarding the racial makeup of individuals visiting emergency departments due to asthma comes from a data set provided by the California Department of Public Health and the California Breathing Asthma Program. The data came from the Office of Statewide Health Planning and Development's emergency department database, which recorded the emergency department visits from all licensed hospitals in California. The designation of a visit due to asthma was based only on primary discharge diagnosis codes. The Age adjusted emergency department visit rate was calculated by dividing the number of visits due to asthma by the estimated population in the county and age group and multiplying by 10,000 to determine the rate per 10,000 people. Age was adjusted to the 2000 census.

## Data Analysis and Statistics

The data sets were loaded into R Studio using the package "readr" and analyzed using several packages. Data was organized using "deplyr" and "tidyr", then plotted graphically using "ggplot2" or spatially using "ggmap". Statistical significance of results were determined with regression analysis using a linear model.

# Results

## Pollution Burden and Asthma Emergency Department Visit Rates by County

# Discussion

# Figures
```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
library("ggmap")
```

```{r read-in-data}
#https://data.ca.gov/dataset/calenviroscreen-3-0-results
enviroscreen_data <- read_csv(
  "data/raw_data/calenviroscreen-3.0-results-june-2018-update.csv")
asthma_rates <- read_csv(
  "data/raw_data/asthma-emergency-department-visit-rates-by-county.csv")
```
## Figure 1
```{r pollution-and-asthma}
#Look at relationship between asthma rates and mean
#pollution burden scores in california counties
#cal enviroscreen: "Pollution Burden scores for each census tract are derived
#from the average percentiles of
#the seven Exposures indicators (ozone and PM2.5 concentrations,
#diesel PM emissions, drinking water contaminants, pesticide use, toxic
#releases from facilities, and traffic density) and the five Environmental
#Effects indicators (cleanup sites, impaired water bodies,groundwater
#threats, hazardous waste facilities and generators, and solid waste sites
#and facilities)."
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE),
            sum_population = sum(`Total Population`)) %>%
  ggplot(aes(x = mean_pollution,
             y = mean_asthma_rate,
             size = sum_population)) +
  geom_point() +
  ggtitle("Relatioship Between Pollution Burden and Asthma Related ED Visits",
         "In California Counties") +
  labs(x =
         "Mean Pollution Burden Score (By County)",
       y = "Mean Rate (per 10,000) of ED Visits Due to Asthma") +
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")
```

```{r statisitcal-analysis-of-pollution-plot}
#Im going to figure out whether this relationship is statistically
#significant
#linear regression fit for pollution burden score and asthma
summarized_pollution_data <- enviroscreen_data %>%
    group_by(`California County`) %>%
  summarize(sum_population = sum(`Total Population`),
            mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE))
#statistical test
summary(lm(mean_asthma_rate ~ mean_pollution,
           data = summarized_pollution_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r organized-PM25-asthma-plot}
#goal is to compare asthma related emergency department visits to mean PM
#2.5 concentration
#Im making a plot to compare the two, using enviroscreen data from 2012-2014
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_pm25,
             y = mean_asthma,
             size = sum_pop)) +
  geom_point() +
  ggtitle("Relationship Between PM2.5 Concentration",
          "and Asthma in California Counties") +
  labs(x = "Mean PM2.5 Concentration",
       y = "Age Adjusted Rate of ED Visits Due to Asthma") +
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")
```

```{r statistical-analysis-of-pm25-and-asthma}
#im going to see if there is a significant relationship between PM2.5
#concentration and rate of asthma related ED visits

#linear regression fit for mean pm2.5 and mean asthma ed visit rates
summarized_pm25_data <- enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE))

#statistical test
summary(lm(mean_asthma ~ mean_pm25, data = summarized_pm25_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r enviroscreen-map}
#since the enviroscreen gives me latitude and longitude, i can make
#map that compares the pm2.5 pollution and asthma ed visit rate in the cesus
#tracts
#filter data from one county
city_data <- filter(enviroscreen_data,
                    `California County` == "San Francisco",
                    `Census Tract` != "6075017902")
#load map
cal_map <- get_map("San Francisco, califoria", zoom = 12)
#plot diesel PM and Asthma
ggmap(cal_map) +
  geom_point(aes(x = Longitude,
                 y = Latitude,
                 color = `Diesel PM`,
                 size = Asthma),
             data = city_data) +
  ggtitle("Diesel Fuel Use and Asthma in San Francisco") +
  labs(x = "Longitude",
       y = "Latitude")
```

```{r regression-analysis-asthma-diesel}
#use regression analysis to see statistical signifigance of the correlation
#shown in the figure
summary(lm(`Diesel PM` ~ Asthma, data = city_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 4)
```


```{r poverty-map}
#can i compare asthma and poverty in san francisco?
city_data <- filter(enviroscreen_data, `California County` == "San Francisco",
                    `Census Tract` != "6075017902")
#load map
cal_map <- get_map("San Francisco, califoria", zoom = 12)
#plot diesel PM and Asthma
ggmap(cal_map) +
  geom_point(aes(x = Longitude,
                 y = Latitude,
                 color = Poverty,
                 size = Asthma),
             data = city_data) +
  ggtitle("Poverty and Asthma in San Francisco") +
  labs(x = "Longitude",
       y = "Latitude")
```

```{r regression-analysis-asthma-poverty}
#like before, do the same regression analysis except with poverty as the
#independent variable
summary(lm(Poverty ~ Asthma, data = city_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r race-asthma-breakdown}
#i would like to observe the relationship between race and asthma in all of the
#california counties
#summarise data
asthma_summary <- asthma_rates %>%
  filter(STRATA == "Race/ethnicity",
         `AGE GROUP` == "All ages",
         YEAR == "2017") %>%
  group_by(`STRATA NAME`) %>%
  summarise(mean_asthma_rate = mean(`AGE-ADJUSTED ED VISIT RATE`,
                            na.rm = TRUE)) %>%
  arrange(desc(mean_asthma_rate))
View(asthma_summary)
#plot graph
ggplot(data = asthma_summary, aes(x = `STRATA NAME`, y = mean_asthma_rate)) +
  geom_col(fill = "cornflowerblue") +
  ggtitle(
    "Emergency Department Visit Rates by Race",
    "Across California Counties") +
  labs(x = "Race/Ethnicity", y = "Mean Asthma Rate (Age Adjusted)")
```

# Sources Cited

