---
title: "Investigating the Relationship Between Asthma and Pollution Burden in California to Understand the Importance of Race and Economic Status as Social Determinants of Health"
author: "Maggie Walsh"
date: "March 2, 2020"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, error = FALSE, working = FALSE)
```

# Introduction
Air pollution has become an increasingly important topic of research in the past half-century, mainly because of the multitude of health risks that have been shown to accompany exposure to air pollution such as respiratory disease, cardiovascular disease, and even neurological dysfunction[@brunekreef2002air; @rajagopalan2018air; @schikowski2020role]. Air pollution consists of a multitude of substances from a multitude of sources. Air pollution arises from fuel burning for energy and transportation, and as a product of other industrial processes. Pollution from mobile sources such as motor traffic has been cited as one of the most prevalent sources of harmful air pollution, especially in cities, where a vast population uses cars to commute into and out of dense urban areas daily[@mayer1999air].

Asthma is a respiratory disease that has been linked to air pollution in numerous studies worldwide[@koenig1999air; @raji2020acute]. There are studies that look specifically at air pollution from mobile sources on the development and worsening of asthma [@gasana2012motor; @ristovski2012respiratory]. These studies investigate a number of different components of motor traffic pollution which contribute to asthma including particulate matter. Particulate matter less than 2.5 micrometers in diameter, known as fine particulate matter, can have a hazardous effect because it is small enough to be inhaled into small airways and alveoli (small air sacs) in the lung[@guarnieri2014outdoor]. These fine particles consist of transition metals and other substances that can cause the symptoms associated with asthma attacks[@guarnieri2014outdoor]. These pollutants, which emerge from cars on major roadways in dangerous concentrations, can have these effects up to 500 meters away[@guarnieri2014outdoor]. For reference, that is the distance covered by five and a half football fields. In North American cities, 30-45% of people live within this distance of a major roadway[@guarnieri2014outdoor].

Air pollution from motor traffic has a particularly pronounced impact in the state of California. In the American Lung Association's State of the Air Report for 2019, 6 out of 10 of the countries most polluted cities for year round particle pollution were California cities. California relies heavily on car commuting because of the development of cities and surrounding infrastructure in a way that prioritized travel by motor vehicles[@cervero2003growing]. The problem is currently worsening because rapidly growing urban areas in Southern California and the Bay Area promise new jobs, but no affordable housing opportunities[@cervero2003growing].Because of the disparity between job opportunities and affordable housing, Californians are forced to drive long distances to work and thereby increase the mobile source pollution in these metropolitan areas. This may contribute to the placement of the Los Angeles and San Jose/San Francisco metropolitan areas on the top ten list for most particle polluted cities. Much of this research has been focused on Southern California, with comparatively little research focused on the Bay Area[@houston2004structural; @houston2014disparities; @perez2009global].

The problem of air pollution and asthma does not impact all Californians equally. Differences in race and socioeconomic status has been shown to relate to how much particle pollution from motor traffic one is exposed to[@houston2004structural; @houston2014disparities]. Race and socioeconomic status, as they relate to potential health impacts, have come to be known as social determinants of health[@marmot2005social]. These social determinants of health reflect how racism and class-ism extend beyond individual viewpoints, into the fabric of our society.

In this study, I aimed to investigate this relationship between asthma and traffic pollution in California Counties using asthma emergency department visit rates and other factors considered by the California Enviroscreen. Then, I questioned the relationship between poverty, race and Asthma in San Francisco. I identified a strong relationship between asthma and poverty in the San Francisco region, and a striking relationship between race and asthma state wide.

# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 



# Discussion


# Figures
```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
library("tidycensus")
library("ggmap")

#tidycensus API key
census_api_key("f26cc4781be4aec1551e0b17e1f052214e3f5256")
```

```{r read-in-data}
#https://data.ca.gov/dataset/calenviroscreen-3-0-results
enviroscreen_data <- read_csv(
  "data/raw_data/calenviroscreen-3.0-results-june-2018-update.csv")
asthma_rates <- read_csv(
  "data/raw_data/asthma-emergency-department-visit-rates-by-county.csv")
```
```{r pollution-and-asthma, message=TRUE, warning=TRUE}
#Look at relationship between asthma rates and mean
#pollution burden scores in california counties
#cal enviroscreen: "Pollution Burden scores for each census tract are derived
#from the average percentiles of
#the seven Exposures indicators (ozone and PM2.5 concentrations,
#diesel PM emissions, drinking water contaminants, pesticide use, toxic
#releases from facilities, and traffic density) and the five Environmental
#Effects indicators (cleanup sites, impaired water bodies,groundwater
#threats, hazardous waste facilities and generators, and solid waste sites
#and facilities)."
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE),
            sum_population = sum(`Total Population`)) %>%
  ggplot(aes(x = mean_pollution,
             y = mean_asthma_rate,
             size = sum_population)) +
  geom_point() +
  ggtitle("Relatioship Between Pollution Burden and Asthma Related ED Visits",
         "In California Counties") +
  labs(x =
         "Mean Pollution Burden Score (By County)",
       y = "Mean Rate (per 10,000) of ED Visits Due to Asthma") +
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")
```

```{r statisitcal-analysis-of-pollution-plot}
#Im going to figure out whether this relationship is statistically
#significant
#linear regression fit for pollution burden score and asthma
summarized_pollution_data <- enviroscreen_data %>%
    group_by(`California County`) %>%
  summarize(sum_population = sum(`Total Population`),
            mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE))
#statistical test
summary(lm(mean_asthma_rate ~ mean_pollution,
           data = summarized_pollution_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r organize-into-plot}
#goal is to compare asthma related emergency department visits to mean PM
#2.5 concentration
#Im making a plot to compare the two, using enviroscreen data from 2012-2014
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_pm25,
             y = mean_asthma,
             size = sum_pop)) +
  geom_point() +
  ggtitle("Relationship Between PM2.5 Concentration",
          "and Asthma in California Counties") +
  labs(x = "Mean PM2.5 Concentration",
       y = "Age Adjusted Rate of ED Visits Due to Asthma") +
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")
```

```{r statistical-analysis-of-pm25-and-asthma}
#im going to see if there is a significant relationship between PM2.5
#concentration and rate of asthma related ED visits

#linear regression fit for mean pm2.5 and mean asthma ed visit rates
summarized_pm25_data <- enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE))

#statistical test
summary(lm(mean_asthma ~ mean_pm25, data = summarized_pm25_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r enviroscreen-map}
#since the enviroscreen gives me latitude and longitude, i can make
#map that compares the pm2.5 pollution and asthma ed visit rate in the cesus
#tracts
#filter data from one county
city_data <- filter(enviroscreen_data,
                    `California County` == "San Francisco",
                    `Census Tract` != "6075017902")
#load map
cal_map <- get_map("San Francisco, califoria", zoom = 12)
#plot diesel PM and Asthma
ggmap(cal_map) +
  geom_point(aes(x = Longitude,
                 y = Latitude,
                 color = `Diesel PM`,
                 size = Asthma),
             data = city_data) +
  ggtitle("Diesel Fuel Use and Asthma in San Francisco") +
  labs(x = "Longitude",
       y = "Latitude")
```

```{r regression-analysis-asthma-diesel}
#use regression analysis to see statistical signifigance of the correlation
#shown in the figure
summary(lm(`Diesel PM` ~ Asthma, data = city_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 4)
```


```{r poverty-map}
#can i compare asthma and poverty in san francisco?
city_data <- filter(enviroscreen_data, `California County` == "San Francisco",
                    `Census Tract` != "6075017902")
#load map
cal_map <- get_map("San Francisco, califoria", zoom = 12)
#plot diesel PM and Asthma
ggmap(cal_map) +
  geom_point(aes(x = Longitude,
                 y = Latitude,
                 color = Poverty,
                 size = Asthma),
             data = city_data) +
  ggtitle("Poverty and Asthma in San Francisco") +
  labs(x = "Longitude",
       y = "Latitude")
```

```{r regression-analysis-asthma-poverty}
#like before, do the same regression analysis except with poverty as the
#independent variable
summary(lm(Poverty ~ Asthma, data = city_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
```

```{r race-asthma-breakdown}
#i would like to observe the relationship between race and asthma in all of the
#california counties
#summarise data
asthma_summary <- asthma_rates %>%
  filter(STRATA == "Race/ethnicity",
         `AGE GROUP` == "All ages",
         YEAR == "2017") %>%
  group_by(`STRATA NAME`) %>%
  summarise(mean_asthma_rate = mean(`AGE-ADJUSTED ED VISIT RATE`,
                            na.rm = TRUE)) %>%
  arrange(desc(mean_asthma_rate))
View(asthma_summary)
#plot graph
ggplot(data = asthma_summary, aes(x = `STRATA NAME`, y = mean_asthma_rate)) +
  geom_col(fill = "cornflowerblue") +
  ggtitle(
    "Emergency Department Visit Rates by Race",
    "Across California Counties") +
  labs(x = "Race/Ethnicity", y = "Mean Asthma Rate (Age Adjusted)")
```

# Sources Cited

