---
title: "Assessment of Hazardous Air Pollution in San Francisco Neighborhoods: Identifying Major Sources"
author: "Maggie Walsh"
date: "March 2, 2020"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, error = FALSE, working = FALSE)
```

# Introduction



# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

```{r read-in-data}
#https://data.chhs.ca.gov
asthma_ed_data <- read_csv(
  "data/raw_data/asthma-emergency-department-visit-rates-by-county.csv")
#https://data.ca.gov/dataset/calenviroscreen-3-0-results
enviroscreen_data <- read_csv(
  "data/raw_data/calenviroscreen-3.0-results-june-2018-update.csv")
```

```{r organize data}
#Look at relationship between asthma rates and traffic density
enviroscreen_data%>%
  group_by(`California County`) %>%
  summarize(mean_traffic_density = mean(Traffic, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE) )%>%
  ggplot(aes(x = mean_traffic_density, 
             y = mean_asthma_rate)) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r organize-into-plot}
#goal is to compare asthma related emergency department visits to mean PM
#2.5 concentration
#Im making a plot to compare the two, using enviroscreed data from 2012-2014
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_pm25, 
             y = mean_asthma, 
             size = sum_pop))+
  geom_point()+
  ggtitle("Relationship Between PM2.5 Concentration", 
          "and Asthma in California Counties")+
  labs(x = "Mean PM2.5 Concentration",
       y = "Age Adjusted Rate of ED Visits Due to Asthma")+
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")

  
```
```{r statistical-analysis-of-pm25-and-asthma}
#im going to see if there is a significant relationship between PM2.5 
#concentration and rate of asthma related ED visits

summarized_county_data <- enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE))

summary(lm(mean_asthma ~ mean_pm25, data = summarized_county_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
        
```

# Discussion

# Sources Cited

