---
title: "Assessment of Hazardous Air Pollution in San Francisco Neighborhoods: Identifying Major Sources"
author: "Maggie Walsh"
date: "March 2, 2020"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, error = FALSE, working = FALSE)
```

# Introduction



# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
library("tidycensus")
library("ggmap")

#tidycensus API key
census_api_key("f26cc4781be4aec1551e0b17e1f052214e3f5256")

```

```{r read-in-data}
#https://data.ca.gov/dataset/calenviroscreen-3-0-results
enviroscreen_data <- read_csv(
  "data/raw_data/calenviroscreen-3.0-results-june-2018-update.csv")
#https://arb.ca.gov/emfac/2014
marin_county_traffic_data <- read_csv(
  "data/raw_data/marin_county_traffic_data.csv")
kings_county_traffic_data <- read_csv(
  "data/raw_data/kings_county_traffic_data.csv")
```


```{r organize data}
#Look at relationship between asthma rates and traffic density
enviroscreen_data%>%
  group_by(`California County`) %>%
  summarize(mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE), 
            sum_population = sum(`Total Population`))%>%
  ggplot(aes(x = mean_pollution, 
             y = mean_asthma_rate, 
             size = sum_population)) +
  geom_point() +
  ggtitle("Relatioship Between Pollution Burden and Asthma Related ED Visits", 
         "In California Counties") +
  labs(x = 
         "Mean Pollution Burden Score given to California Census Tracts (By County)", 
       y = "Mean Rate (per 10,000) of ED Visits Due to Asthma") +
  geom_smooth(method = lm) 
```

```{r statisitcal-analysis-of-pollution-plot}
#Im going to figure out whether this relationship is statistically significant
#linear regression fit for pollution burden score and asthma
summarized_pollution_data <- enviroscreen_data %>%
    group_by(`California County`) %>%
  summarize(sum_population = sum(`Total Population`), 
            mean_pollution = mean(`Pollution Burden Score`, na.rm = TRUE),
            mean_asthma_rate = mean(Asthma, na.rm = TRUE))
            
#statistical test
summary(lm(mean_asthma_rate ~ mean_pollution, 
           data = summarized_pollution_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)  
```

```{r organize-into-plot}
#goal is to compare asthma related emergency department visits to mean PM
#2.5 concentration
#Im making a plot to compare the two, using enviroscreen data from 2012-2014
enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_pm25, 
             y = mean_asthma, 
             size = sum_pop))+
  geom_point()+
  ggtitle("Relationship Between PM2.5 Concentration", 
          "and Asthma in California Counties")+
  labs(x = "Mean PM2.5 Concentration",
       y = "Age Adjusted Rate of ED Visits Due to Asthma")+
  geom_smooth(method = lm) +
  scale_size_continuous(name = "County Population Size")
  
```

```{r statistical-analysis-of-pm25-and-asthma}
#im going to see if there is a significant relationship between PM2.5 
#concentration and rate of asthma related ED visits

#linear regression fit for mean pm2.5 and mean asthma ed visit rates
summarized_pm25_data <- enviroscreen_data %>%
  group_by(`California County`) %>%
  summarize(sum_pop = sum(`Total Population`),
            mean_pm25 = mean(PM2.5, na.rm = TRUE),
            mean_asthma = mean(Asthma, na.rm = TRUE))

#statistical test
summary(lm(mean_asthma ~ mean_pm25, data = summarized_pm25_data)) %>%
  broom::tidy() %>%
  knitr::kable(digits = 3)
        
```

```{r enviroscreen-map}
#since the enviroscreen gives me latitude and longitude, i can make a 
#map that compares the pm2.5 pollution and asthma ed visit rate in the cesus 
#tracts
#filter data from one county
city_data <- filter(enviroscreen_data, `California County` == "San Francisco") 
#load map
cal_map <- get_map("San Francisco, califoria", zoom = 12)
#plot diesel PM and Asthma
ggmap(cal_map) +
  geom_point(aes(x = Longitude, 
                 y = Latitude, 
                 color = `Diesel PM`,
                 size = Asthma),
             data = city_data) +
  ggtitle("Diesel Fuel Use and Asthma in San Francisco") +
  labs(x = "Longitude",
       y = "Latitude")


```



# Discussion

# Sources Cited

